---
title: "Analyzing the 2016 US Presidential Election"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

We analyze returns from the 2012 and 2016 elections in order to
understand the social and demographic trends that may have contributed
to Donald Trump's victory in 2016. We will first examine how Republican
vote share at the county level has changed from 2012 to 2016. Then, we
will look at four variables that were prominent in the discourse around
the election -- race, education, unemployment, and immigration -- to see
how well they predict GOP electoral gains at the county level.

We will be working with three datasets. The first, `election2012.csv`,
has one observation per county and contains the following variables:

| Name             | Description                                                  |
|:-----------------|:-----------------------------------------------------|
| `FIPS`           | FIPS code (unique county identifier)                         |
| `state`          | State abbreviation                                           |
| `county`         | County name                                                  |
| `votes_dem_12`   | Number of votes cast for Democratic candidate, 2012 election |
| `votes_gop_12`   | Number of votes cast for Republican candidate, 2012 election |
| `votes_total_12` | Total number of votes cast in 2012 election                  |

The second, `election2016.csv`, has the same data structure and similar
variable names but reports data for the 2016 presidential election.

The third dataset, `county.csv`, includes social and demographic
characteristics for each county:

| Name              | Description                                                                                                                                                     |
|:-----------------|:-----------------------------------------------------|
| `FIPS`            | FIPS code (unique county identifier)                                                                                                                            |
| `pct_for_born15`  | Percent of county's population that is "foreign born" according to the U.S. Census, meaning anyone who is not a U.S. citizen at birth (measured over 2011-2015) |
| `pct_bach_deg15`  | Percent of county population holding a Bachelor's degree or above (2011-2015)                                                                                   |
| `pct_non_white15` | Percent of county population that is not white (2011-2015)                                                                                                      |
| `pct_unemp16`     | Percent of county population that is unemployed, BLS estimates (average, Jan-Oct 2016)                                                                          |
| `pct_unemp12`     | Percent of county population that is unemployed, BLS estimates (average, Jan-Oct 2012)                                                                          |

## Question 1

Start by load all three datasets. Merge the three datasets by FIPS code
to construct one complete data file for analysis. Check your merge to
see how many observations came in from all three sources. Did you lose
much data? Finally, perform listwise deletion (hint: check section 3.2
of *QSS*) of missing values on the full dataset. Did you lose much data?
Get whatever characteristics you can on the data you lost. What can you
say about these observations?

## Answer 1

```{r}
## read in data
returns12 <- read.csv("/Users/juansantiagobarragan/Documents/GitHub/first_unguided_project-team_2/election2012.csv")
returns16 <- read.csv("/Users/juansantiagobarragan/Documents/GitHub/first_unguided_project-team_2/election2016.csv")
covars <- read.csv("/Users/juansantiagobarragan/Documents/GitHub/first_unguided_project-team_2/county.csv")

## which columns do we want in both? No need to duplicate state and county.
#FIPS votes_dem_12 votes_gop_12 votes_total_12 votes_dem_16 votes_gop_16 votes_total_16 pct_for_born15 pct_bach_deg15 pct_non_white15 pct_unemp12 pct_unemp16
## now, lets merge
p1merge <- merge(returns12,returns16[c("FIPS","votes_dem_16","votes_gop_16","votes_total_16")],by="FIPS")

# merged datasets for votes for 2012 and 2016, made sure to merge columns that was not 'County' or 'State'
## did we lose any observations?
#No Observations were lost, 3141 observations remain 
## now let's merge on covariates

county_merge <- merge(p1merge,covars,by="FIPS") #merged the covariates that take into account County demographics with p1merge

## remove missing values (listwise deletion)
complete_df <- county_merge[complete.cases(county_merge), ] #removed rows that contain any missing values to clean our dataset 

## how much data did we lose?
#We lost 29 rows 
## identify lost data
county_merge[!complete.cases(county_merge),] #identified rows that had N/A part of the dataset
```

## Question 2

Compute the Republican vote share as a proportion of total votes, in
2012 as well as in 2016. Also compute the percent difference in this
Republican vote share variable from the 2012 to 2016 election. Plot the
distribution of this percent difference, with a red line at the median.

Then, subset your data to just the battleground states: Florida, North
Carolina, Ohio, Pennsylvania, New Hampshire, Michigan, Wisconsin, Iowa,
Nevada, Colorado, and Virginia. Plot the distribution of the same
variable in this sample, with a red line at the sample median.

## Answer 2

```{r}
library(ggplot2)
## vote share variables: Republican / Total
rep_12_voteshare <- county_merge$votes_gop_12/county_merge$votes_total_12
rep_16_voteshare <- county_merge$votes_gop_16/county_merge$votes_total_16
voteshares <- data.frame(rep_12_voteshare,rep_16_voteshare)

## percent change from 2012 to 2016
pct_change <- (rep_16_voteshare - rep_12_voteshare)/rep_12_voteshare #used percent increase formula 

## plot the distributions
pct_df <- data.frame(pct_change)
ggplot(pct_df,aes(x=pct_change)) + geom_histogram() + geom_vline(xintercept = median(median(pct_df$pct_change,na.rm=TRUE)),col = "red") + ylab("Numbner of Counties") + xlab("Percent Change for all Counties") 

#made a histogram to accurately show distribution for how much the vote percent difference was per county. Utilizing ggplot, we called 'pct_df' as the dataset and used the column 'pct_change' to the x-axis. We then used geom_histogram() to make it a histogram and used geom_vline() to create the red line to represnt the median line. Finally, we used labs() to label the x and y axis. 

# This plot shows us that Counties/States barely changed their vote percentage between.Most states are near the 0 percent difference which is not a huge shift in vote choices. 

## subset data to battleground states
battlestates <- c("FL", "NC", "OH", "PA", "NH", "MI", "WI", "IA", "NV", "CO", "VA")
bg_states <- subset(county_merge, state == c("FL", "NC", "OH", "PA", "NH", "MI", "WI", "IA", "NV", "CO", "VA"))
rep_12_voteshare_bg <- bg_states$votes_gop_12/bg_states$votes_total_12
rep_16_voteshare_bg <- bg_states$votes_gop_16/bg_states$votes_total_16
pct_change_bg <- (rep_16_voteshare_bg - rep_12_voteshare_bg)/rep_12_voteshare_bg
pct_df_bg <- data.frame(pct_change_bg)
ggplot(pct_df_bg,aes(x=pct_change_bg)) + geom_histogram() + geom_vline(xintercept = median(pct_df_bg$pct_change_bg,na.rm=TRUE),col = "red") + ylab("Numbner of Counties") + xlab("Percent Change in Counties located in Battleground States")

#made a histogram to accurately show distribution for how much the vote percent difference was per county for Battleground states. Utilizing ggplot, we called 'pct_df_bg' as the dataset and used the column 'pct_change_bg' to the x-axis. We then used geom_histogram() to make it a histogram and used geom_vline() to create the red line to represnt the median line. Finally, we used labs() to label the x and y axis. 
#This plot showed us that unlike other counties in the United States, the percent change in Battleground States only increased positively meaning that Counties/States further cemented made their political beliefs instead of flipping. There is a peak of Counties/States that barely change their vote percent difference but there is more Counties/States that cemented their political beliefs more. 
```